\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2019-12-11 sections/methods.tex
Updated draft of manuscript

  \vspace{-5mm}
  The study uses three kinds of 16S rRNA sequencing datasets \- real datasets, mock datasets and synthetic datasets.
  Real datasets are those sequencing reads obtained from naturally occurring microbial community samples.
  The current study used healthy stool samples from a fecal microbiome transplant study \cite{Kang2017} and healhty saliva samples from a periodontal disease study~\cite{Chen2018} as real datasets for analysis.
  The mock community 16S datasets were obtained from mockrobiota~\cite{Bokulich2016}.
  The mock datasets used for this study were labelled mock4, mock12 and mock16 datasets.
  The synthetic datasets were generated using an artificial read simulator called ART~\cite{Huang2012}.
  Three different microbial composition profiles were used as input
  Synthetic reads were generated using a soil and water microbiome composition profiles from the Earth Microbiome Project~\cite{Thompson2017} and healthy gut microbiome project from the Human Microbiome Project~\cite{HumanMicrobiomeProjectConsortium2012}
  The scripts used to generate the synthetic data are in the \hl{scripts folder of the respository}

@@

  The mindpipe pipeline provides both a Python API as well as a command-line interface and only requires a single configuration file.
  The configuration file lists the inputs, output and the steps to be performed during runtime, along with the parameters to be used (if different from defaults) for the various steps.
  Since the entire pipeline run-through is stored in the form of a text file (the configuration file), subsequent runs are highly reproducible and changes can be easily tracked using version control.
  It uses the nextflow workflow manager~\cite{Tommaso2015} under the hood and facilitates the usage of the pipeline on the local machine, cluster or cloud with minimal configuration change.
  It also allows for automatic parallelization of all possible processes, both within and across samples.
  The pipeline is designed to be modular, where each tool or method is organized into modules which can be easily modified or replaced.
  This also simplifies the process of adding new tools (refer to modules section in the mindpipe documentation).
  In addition to the Python package, the entire pipeline has been containerized into a Docker~\cite{Merkel1994} image (\hl{dockerhub link}) for easy deployment and setup.
  The main components of the pipeline are detailed in the subsequent sections.

@@

  \subsection*{Network Inference}
  \vspace{-5mm}
  This module deals with the inference of co-occurrence associations from the \ac{otu} or \ac{otu} counts matrix.
  These associations can be represented as a network, with nodes representing taxonomies and edges the association between them.
  \begin{itemize}
    \item A null model is created by re-sampling and bootstrapping the correlation/interaction matrix
    \item Significance of these associations is determined by calculating pvalues against this null model.
  \end{itemize}
  The pipeline includes Pearson, Spearman and \ac{sparcc}~\cite{Friedman2012} as the pairwise correlation metrics, and \ac{spieceasi}~\cite{Kurtz2015}, \ac{mldm}~\cite{Yang2017} and \ac{magma} (\hl{in biorxiv}) as the direct association metrics.

@@

  \textbf{Power calculations:}

  \textbf{P-value merging}
  The empirical Browns method is used for combining p-values from the various methods to obtain a consensus pvalue, which is used to create the consensus network.

@@

  Links here

>>>>

  \vspace{-5mm}
  The study uses three kinds of 16S rRNA sequencing datasets \- real datasets, mock datasets and synthetic datasets.
  Real datasets are those sequencing reads obtained from naturally occurring microbial community samples.
  The current study used healthy stool samples from a fecal microbiome transplant study \cite{Kang2017} and healhty saliva samples from a periodontal disease study~\cite{Chen2018} as real datasets for analysis.
  The mock community 16S datasets were obtained from mockrobiota~\cite{Bokulich2016}.
  The mock datasets used for this study were labelled mock4, mock12 and mock16.
  The mock4 community is composed of 21 bacterial strains represented in equal abundances in two replicate samples, and same strains in unequal abundances in two more replicate samples.
  The mock12 community is composed of 27 bacterial strains that include closely related taxa with some pairs having only one to two nucleotide difference from another.
  The mock16 community is composed of equal amounts of 49 bacteria and 10 archea.
  The synthetic datasets were generated using an artificial read simulator called ART~\cite{Huang2012}.
  Three different microbial composition profiles were used as input
  Synthetic reads were generated using a soil and water microbiome composition profiles from the Earth Microbiome Project~\cite{Thompson2017} and healthy gut microbiome project from the Human Microbiome Project~\cite{HumanMicrobiomeProjectConsortium2012}.
  The scripts used to generate the synthetic data are in the scripts folder of the respository (\href{https://github.com/dileep-kishore/mind-pipeline-paper}{https://github.com/dileep-kishore/mind-pipeline-paper})

@@

  The mindpipe pipeline provides both a Python API as well as a command-line interface and only requires a single configuration file.
  The configuration file lists the inputs, output and the steps to be performed during runtime, along with the parameters to be used (if different from defaults) for the various steps.
  Since the entire pipeline run-through is stored in the form of a text file (the configuration file), subsequent runs are highly reproducible and changes can be easily tracked using version control.
  It uses the nextflow workflow manager~\cite{Tommaso2015} under the hood and facilitates the usage of the pipeline on the local machine, cluster or cloud with minimal configuration change.
  It also allows for automatic parallelization of all possible processes, both within and across samples.
  The pipeline is designed to be modular, where each tool or method is organized into modules which can be easily modified or replaced.
  This also simplifies the process of adding new tools (refer to modules section in the mindpipe documentation).
  In addition to the Python package, the entire pipeline has been containerized into a Docker~\cite{Merkel1994} image (dockerhub link) for easy deployment and setup.
  The main components of the pipeline are detailed in the subsequent sections.

@@

  \subsection*{Network Inference}
  \vspace{-5mm}
  This module deals with the inference of co-occurrence associations from the \ac{otu} or \ac{otu} counts matrix.
  These associations can be represented as a network, with nodes representing taxonomies and edges the association between them.
  \begin{itemize}
    \item A null model is created by re-sampling and bootstrapping the correlation/interaction matrix
    \item Significance of these associations is determined by calculating pvalues against this null model.
  \end{itemize}
  The pipeline includes Pearson, Spearman and \ac{sparcc}~\cite{Friedman2012} as the pairwise correlation metrics, and \ac{spieceasi}~\cite{Kurtz2015}, \ac{mldm}~\cite{Yang2017} and \ac{magma} (in biorxiv) as the direct association metrics.

@@

  % TODO: Variance and PCA calculation

  % \textbf{Power calculations:}

\textbf{P-value merging}:
The empirical Browns method [ref] is used for combining p-values from the various methods to obtain a consensus pvalue, which is used to create the consensus network.

@@

### 2019-12-11 sections/results.tex
Updated draft of manuscript

  \begin{figure}
    \centering
    \includegraphics[width=0.85\linewidth]{figure2.pdf}
    \caption{
      \textbf{The choice of database contributes to the most variance in the networks}.
      \textbf{(A)} The relative variance in the networks contributed by the DC, TA and OP steps of the pipeline.
      \textbf{(B)} All combinations of inferred networks are shown as points on a PCA plot.
      The color of the points corresponds to the taxonomy database, the shape corresponds to the denoising/clustering method and the size corresponds to whether low abundance OTUs were removed or not.
      \textbf{(B inset)} The network generated using DC=dada2, TA=GG, OP=no and NI=SPARCC and represents the particular point shown (big red square).
      The plot clearly shows that the points can be separated based on the TA step and that the differences due to the DC and OP steps are not as significant.
    }
    \label{fig:figure2}
  \end{figure}

@@

  Denoising and clustering are commonly carried out to generate representative sequences from the raw 16S sequencing data and to obtain the \ac{otu}/\ac{esv} tables (counts of these representative sequences for each sample).
  In order to compare the \ac{otu} tables generated by these tools we process the same 16S sequencing reads (\hl{FMT} dataset) using 5 different methods \- open-reference clustering, closed-reference clustering, denovo clustering, \ac{dada2}~\cite{Callahan2016} and Deblur~\cite{Amir2017}.
  The first three methods are from the \ac{qiime1}~\cite{Caporaso2010} package.
  We find that there is good agreement in the \ac{otu}/\ac{esv} tables when different combinations of methods are used to generate them.

@@

  There is no method that clearly outperforms the rest in all datasets.
  Based on their slightly better performance on the mock datasets, their \textit{de novo} error correcting nature and other previous studies [refs] we recommend either \ac{dada2} or Deblur for users.
  However, as a result of the unexpected bad performance of Deblur on the synthetic data, the default algorithm in the pipeline was chosen to be \ac{dada2}.

>>>>

  \begin{figure}
    \centering
    \includegraphics[width=0.85\linewidth]{figure2.pdf}
    \caption{
      \textbf{The choice of database contributes to the most variance in the networks}.
      \textbf{(A)} The total relative variance in the networks contributed by the DC, TA and OP steps of the pipeline (right) and the linear model used to calculate the relative variance (left).
      \textbf{(B)} All combinations of inferred networks are shown as points on a PCA plot.
      The color of the points corresponds to the taxonomy database, the shape corresponds to the denoising/clustering method and the size corresponds to whether low abundance OTUs were removed or not.
      \textbf{(B inset)} The network generated using DC=dada2, TA=GG, OP=no and NI=SPARCC and represents the particular point shown (big red square).
      The plot clearly shows that the points can be separated based on the TA step and that the differences due to the DC and OP steps are not as significant.
    }
    \label{fig:figure2}
  \end{figure}

@@

  Denoising and clustering are commonly carried out to generate representative sequences from the raw 16S sequencing data and to obtain the \ac{otu}/\ac{esv} tables (counts of these representative sequences for each sample).
  In order to compare the \ac{otu} tables generated by these tools we process the same 16S sequencing reads (healthy samples from a fecal microbiome transplant study [ref]) using 5 different methods \- open-reference clustering, closed-reference clustering, denovo clustering, \ac{dada2}~\cite{Callahan2016} and Deblur~\cite{Amir2017}.
  The first three methods are from the \ac{qiime1}~\cite{Caporaso2010} package.
  We find that there is good agreement in the \ac{otu}/\ac{esv} tables when different combinations of methods are used to generate them.

@@

  There is no method that clearly outperforms the rest in all datasets.
  Based on their slightly better performance on the mock datasets, their \textit{de novo} error correcting nature and other previous studies [refs] we recommend either \ac{dada2} or Deblur for users.
  However, as a result of the unexpected poor performance of Deblur on the synthetic data, the default algorithm in the pipeline was chosen to be \ac{dada2}.
